A sheaf--function dictionary for algebraic tori over local fields

### 2013-10-03 shanghai.tex
Edits, mainly to section 7

\begin{proposition}\label{prop:bounded}
 $\QCf(G)$ is (equivalent to) a full subcategory of $\QCb(G)$.
\end{proposition}

@@

\begin{proposition}\label{prop:connected}
  If $G$ is a connected commutative algebraic group over $\Fq$ then 
  the full subcategories $\QCf(G) \hookrightarrow \QCb(G) \hookrightarrow \QC(G)$ 
  appearing in Section~\ref{sec:bounded} are equivalences 
  and
  \[
  \trFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
  is an isomorphism of groups.
\end{proposition}

@@

  Let $g$ be any $\Fq$-rational point on $G$ 
  and let $\bg$ be the geometric point on $G$ lying above $g$ 
  (we fixed $\Fq \hookrightarrow \bFq$ long ago). 
  Since $G$ is connected, the geometric point $\bg$ determines
  an equivalence between the category of $\ell$-adic Weil local systems on $G$ and
  $\ell$-adic representations of $\W(G,\bg)$ \cite{deligne:80a}*{1.1.12}. 
  Now let $(\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$ 
  and let $\rho : \W(G, \bg) \to \EEx$ be the character determined by $(\gqcs{L},\phi)$. 
  The $\Fq$-rational point $g$ under the geometric point $\bg$ determines a splitting
  $\Weil{\Fq}\to \W(G,\bg)$ of $\W(G,\bg)\to \Weil{\Fq}$. 
  In this way the Weil local system $(\gqcs{L},F)$ determines an $\ell$-adic character $\rho_g : \Weil{\Fq} \to \EEx$.
  %\[
  %\begin{tikzcd}
  %1 \rar & \pi_1(\bG, \bg) \rar &  \arrow{d}{\rho} \W(G,\bg) \rar
  %& \ar[swap,bend right, dotted]{l}{g} \arrow[dotted]{dl}{\rho_g} \Weil{\Fq} \rar & 1\\
  %&&\EEx&&
  %\end{tikzcd}
  %\]
  This character is given by the trace of Frobenius of $\qcs{L}$ at $g$, as defined in Section~\ref{sec:Frob}:
  \[
  \rho_g(\Frob{\Fq}) =  \trFrob{\qcs{L}}(g).
  \]

  On the other hand, we have already seen that $\trFrob{\qcs{L}} : G(\Fq) \to \EEx$
  is a group homomorphism. Since $G$ is an algebraic group over $\Fq$, it is a
  variety over $\Fq$, so finitely generated over $\Fq$, so $G(\Fq)$ is finite;
  it follows that $\trFrob{\qcs{L}}(g) = \rho_g(\Frob{\Fq})$ is a root of unity,
  for every $g\in G(\Fq)$.  Since $\Weil{\Fq}$ is generated by
  $\Frob{\Fq}$ and $\rho_g : \Weil{\Fq} \to \EEx$ is
  a character, it follows that the image of $\rho_g$ is a finite group.
  Thus, $\rho_g$ extends to an $\ell$-adic character of $\Gal(\bFq/\Fq)$, 
  denoted by the same symbol, below.
  \[
  \begin{tikzcd}
  1 \rar & \ar[equal]{d} \pi_1(\bG, \bg) \rar & \W(G,\bg) \rar \dar & \Weil{\Fq} \arrow[bend left]{rr}{\rho_g} \rar \dar & 1 & \EEx\\
  1 \rar &  \pi_1(\bG, \bg) \rar & \pi_1(G,\bg) \rar & \Gal(\bFq/\Fq) \arrow[dotted]{rru} \rar & 1 &
  \end{tikzcd}
  \]

@@

  To show that $L_G(f,\psi) \iso (\qcs{L}_0,\mu_0)$ we simply apply the functions--sheaves dictionary
  as follows ({\it cf} \cite{laumon:87a}*{1.1.3} or, ultimately, \cite{deligne:SGA4.5}*{Sommes trigonométriques}).
  First, recall that $L_G(f,\psi) \ceq (L(f,\psi), \mu(f,\psi))$.
  Let $k'/k$ be any degree-$n$ extension and let $\Nm : k'\to k$ be the norm map. 
  By \cite{laumon:87a}*{1.1.3.3}, 
  \[
  \trFrob{(G_{k'}\hookrightarrow G)^*L(f,\psi)} = \trFrob{\qcs{L}} \circ \Nm;
  \]
  on the other hand, 
  \[
  \trFrob{\qcs{L}} \circ \Nm = \trFrob{(G_{k'}\hookrightarrow G)^*\qcs{L}},
  \]
   by Proposition~\ref{prop:basechange}.
  %Using notation from \cite{laumon:87a}*{1.1}, this means $t_{\bullet L(f,\psi)} = t_{\bullet \qcs{L}_0}$.
  Using \cite{laumon:87a}*{Th\'eor\`eme~1.1.2}, it follows that  $L(f,\psi) \iso \qcs{L}_0$ in $\Loc(G)$.
  It is now clear that $(L(f,\psi),\mu(f,\psi)) \iso (\qcs{L}_0,\mu_0)$ in $\QCb(G)$.
  This completes the proof that $L_G : \QCf(G) \to \QC(G)$ is essentially surjective.
  Since we have already seen (in Section~\ref{sec:finite}) that $L_G$ is full and faithful, 
  it follows that $L_G$ is an equivalence.

@@

As Proposition~\ref{prop:connected} shows, 
when $G$ is a connected algebraic group over $\Fq$, 
it is appropriate to replace quasicharacter sheaves on $G$ 
with the conceptually simpler category $\QCf(G)$ of finite quasicharacter sheaves on $G$. 
In this context, finite quasicharacter sheaves may also be apprehended as
\emph{certain} $\ell$-adic characters of fundamental group $\pi_1(G,{\bar e})$. 
As explained in \cite{kamgarpour:09a}*{\S~2}, 
the relevant characters of $\pi_1(G,{\bar e})$ are precisely those 
that factor through a particular quotient of $\pi_1(G,{\bar e})$ 
denoted there by $\Pi_\text{disc}(G)$.
%\[\pi_1(G,{\bar e}) \to \Pi_\text{et}(G) \to \Pi_\text{et}(G)_{\Gal(\bFq/\Fq)} = \Pi_\text{disc}(G)\]
From this it is apparent that the category of discrete isogenies to $G$ is a Galois category,
in the sense of \cite{grothendieck:SGA1}*{Expos\'e V, \S 4}. Moreover, since $\Pi_\text{disc}(G) \iso G(\Fq)$,
as shown in \cite{kamgarpour:09a}*{App'x B}, this provides an alternate proof of
$\QCiso{G} \iso G(\Fq)^*$ when $G$ is a connected algebraic group over $\Fq$.

@@

\'Etale group schemes form a counterpoint to connected groups,
since the component group of a $\QC$-scheme
is an \'etale group scheme, locally of finite type over $\Fq$
\cite{vdGeer-Moonen:AbelianVarieties}*{III, \S 4}.
The functor $G \mapsto G(\bFq)$ defines an equivalence of categories
between the category of \'etale group schemes over $\Fq$ and the category of groups equipped
with an action of $\Gal(\bFq/\Fq)$, continuous for the discrete topology on the group.
Under this equivalence, a quasicharacter sheaf on $G$ is given by the data of:
\begin{enumerate}
 \labitem{(cs.0)}{cs.0} an indexed set of one-dimensional
  $\EE$-vector spaces $\gqcs{L}_x$, as $x$ runs over
  $G(\bFq)$;

>>>>

\begin{proposition}\label{prop:bounded}
 $\QCf(G)$ is equivalent to a full subcategory of $\QCb(G)$.
\end{proposition}

@@

\begin{proposition}\label{prop:connected}
  If $G$ is a connected commutative algebraic group over $\Fq$ then 
  the full subcategories $\QCf(G) \hookrightarrow \QCb(G) \hookrightarrow \QC(G)$ 
  appearing in Sections~\ref{sec:bounded} and \ref{sec:finite} are equivalences 
  and
  \[
  \trFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
  is an isomorphism of groups.
\end{proposition}

@@

  Let $g$ be any $\Fq$-rational point on $G$ 
  and let $\bg$ be the geometric point on $G$ lying above $g$ 
  Since $G$ is connected, the geometric point $\bg$ determines
  an equivalence between the category of $\ell$-adic Weil local systems on $G$ and
  $\ell$-adic representations of $\W(G,\bg)$ \cite{deligne:80a}*{1.1.12}.  \David{We should include a definition for $\W(G,\bg)$, perhaps in an intro paragraph for the section}
  Now let $(\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$ 
  and let $\rho : \W(G, \bg) \to \EEx$ be the character determined by $(\gqcs{L},\phi)$. 
  The $\Fq$-rational point $g$ under the geometric point $\bg$ determines a splitting
  $\Weil{\Fq}\to \W(G,\bg)$ of $\W(G,\bg)\to \Weil{\Fq}$. 
  In this way the Weil local system $(\gqcs{L},\phi)$ determines an $\ell$-adic character $\rho_g : \Weil{\Fq} \to \EEx$.
  %\[
  %\begin{tikzcd}
  %1 \rar & \pi_1(\bG, \bg) \rar &  \arrow{d}{\rho} \W(G,\bg) \rar
  %& \ar[swap,bend right, dotted]{l}{g} \arrow[dotted]{dl}{\rho_g} \Weil{\Fq} \rar & 1\\
  %&&\EEx&&
  %\end{tikzcd}
  %\]
  This character is given by the trace of Frobenius of $\qcs{L}$ at $g$, as defined in Section~\ref{sec:Frob}:
  \[
  \rho_g(\Frob{\Fq}) =  \trFrob{\qcs{L}}(g).
  \]
  \David{It's not clear to me whether we've just proven this claim or if we're citing it from somewhere}

  On the other hand, we have already seen that $\trFrob{\qcs{L}} : G(\Fq) \to \EEx$
  is a group homomorphism. Since $G$ is an algebraic group over $\Fq$, it is a
  variety over $\Fq$ and thus $G(\Fq)$ is finite.
  Therefore $\trFrob{\qcs{L}}(g) = \rho_g(\Frob{\Fq})$ is a root of unity
  for every $g\in G(\Fq)$.  Since $\Weil{\Fq}$ is generated by
  $\Frob{\Fq}$ and $\rho_g : \Weil{\Fq} \to \EEx$ is
  a character, it follows that the image of $\rho_g$ is a finite group.
  Thus, $\rho_g$ extends to an $\ell$-adic character of $\Gal(\bFq/\Fq)$, 
  denoted by the same symbol, below.
  \[
  \begin{tikzcd}
  1 \rar & \ar[equal]{d} \pi_1(\bG, \bg) \rar & \W(G,\bg) \rar \dar & \Weil{\Fq} \arrow[bend left]{rr}{\rho_g} \rar \dar & 1 & \EEx\\
  1 \rar &  \pi_1(\bG, \bg) \rar & \pi_1(G,\bg) \rar & \Gal(\bFq/\Fq) \arrow[dotted]{rru} \rar & 1 &
  \end{tikzcd}
  \]

@@

  To show that $L_G(f,\psi) \iso (\qcs{L}_0,\mu_0)$ we simply apply the sheaf--function dictionary
  as follows \citelist{\cite{laumon:87a}*{1.1.3} \cite{deligne:SGA4.5}*{Sommes trigonométriques}}.
  First, recall that $L_G(f,\psi) \ceq (L(f,\psi), \mu(f,\psi))$.
  Let $k'/k$ be any degree-$n$ extension and let $\Nm : k'\to k$ be the norm map. 
  By \cite{laumon:87a}*{1.1.3.3}, 
  \[
  \trFrob{(G_{k'}\hookrightarrow G)^*L(f,\psi)} = \trFrob{\qcs{L}} \circ \Nm;
  \]
  on the other hand, 
  \[
  \trFrob{\qcs{L}} \circ \Nm = \trFrob{(G_{k'}\hookrightarrow G)^*\qcs{L}},
  \]
   by Proposition~\ref{prop:basechange}.
  %Using notation from \cite{laumon:87a}*{1.1}, this means $t_{\bullet L(f,\psi)} = t_{\bullet \qcs{L}_0}$.
  Using \cite{laumon:87a}*{Th\'eor\`eme~1.1.2}, it follows that  $L(f,\psi) \iso \qcs{L}_0$ in $\Loc(G)$.
  It is now clear that $(L(f,\psi),\mu(f,\psi)) \iso (\qcs{L}_0,\mu_0)$ in $\QCb(G)$.
  This completes the proof that $L_G : \QCf(G) \to \QC(G)$ is essentially surjective.
  Since we have already seen (in Section~\ref{sec:finite}) that $L_G$ is full and faithful, 
  it follows that $L_G$ is an equivalence.

@@

As Proposition~\ref{prop:connected} shows, 
when $G$ is a connected algebraic group over $\Fq$, 
it is appropriate to replace quasicharacter sheaves on $G$ 
with the conceptually simpler category $\QCf(G)$ of finite quasicharacter sheaves on $G$. 
In this context, finite quasicharacter sheaves may also be apprehended as
\emph{certain} $\ell$-adic characters of fundamental group $\pi_1(G,{\bar e})$. 
As explained in \cite{kamgarpour:09a}*{\S~2}, 
the relevant characters of $\pi_1(G,{\bar e})$ are precisely those 
that factor through a particular quotient of $\pi_1(G,{\bar e})$ 
denoted there by $\Pi_\text{disc}(G)$.
%\[\pi_1(G,{\bar e}) \to \Pi_\text{et}(G) \to \Pi_\text{et}(G)_{\Gal(\bFq/\Fq)} = \Pi_\text{disc}(G)\]
This discussion demonstrates that the category of discrete isogenies to $G$ is a Galois category
\cite{grothendieck:SGA1}*{Expos\'e V, \S 4}. Moreover, since $\Pi_\text{disc}(G) \iso G(\Fq)$
\cite{kamgarpour:09a}*{App'x B}, this \David{What does ``this'' refer to} provides an alternate proof that
$\QCiso{G} \iso G(\Fq)^*$ when $G$ is a connected algebraic group over $\Fq$.

@@

\'Etale group schemes form a counterpoint to connected groups,
since the component group of a $\QC$-scheme
is an \'etale group scheme \cite{vdGeer-Moonen:AbelianVarieties}*{III, \S 4}.
The functor $G \mapsto G(\bFq)$ defines an equivalence of categories
between the category of \'etale group schemes over $\Fq$ and the category of groups equipped
with an action of $\Gal(\bFq/\Fq)$, continuous for the discrete topology on the group.
Under this equivalence, a quasicharacter sheaf on $G$ is given by the data of:
\begin{enumerate}
 \labitem{(cs.0)}{cs.0} an indexed set of one-dimensional
  $\EE$-vector spaces $\gqcs{L}_x$, as $x$ runs over
  $G(\bFq)$;
